### 2020-08-31 thesis/thesis.tex
[thesis] Mention value derivation available in repo
[thesis] Corrections, spelling

Algebraic effects, while not a new concept, have received a lot of
attention~\cite{effects-bibliography} in recent years, both from the theoretical
and practical side.

They give the developer necessary tools for declarative programming with
computational effects, on a high level of abstraction, by defining
effects as an API\@ in the code. At the same time, algebraic effects preserve
a lot of flexibility and have a strong theoretical foundations.

@@

In particular, multi-shot resumptions in effect handlers and the possibility to
drop the resumption allow one to express fairly complex logic in a concise way,
but at the same time, they give rise to issues that would not occur in standard
control flow, especially when composing various effects together. Unwanted
interaction with external resources multiple times or not invoking finalisation
code are cases that may occur while using algebraic effects and handlers. Let's
consider the following example:

\vbox{\begin{verbatim}
    handle
        let fh <- do Open "praise.txt" in
        let c <- do Choice () in
        let text <- if c then return "Guy Fieri" else return "is cool" in
        let _ <- do Write (fh, text) in
        do Close fh
    with {
        Choose _ k ->
            let t <- r 1 in
            let f <- r 0 in
            return (t, f) |
        return x -> return x
    }
\end{verbatim}
}

@@

Other branch of languages theory, that we shall investigate more in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed for various use cases, most notably, type systems. Thanks to strong
and static type systems along with their implementations, we have solid tools to
work efficiently on functions that are pure. That being said, we claim that the
core complexity of programs comes from side effects, or more generally, computational
effects, which we cannot avoid in writing anything useful.

@@

\noindent
TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource

Not only we would like to bring back effects to our type
system~\cite{type-and-effect}, but also do it in a way that is composable. This
is where algebraic effects come to the rescue. From theoretical point of view,
we need to develop equational theory about our effects to assert correctness of
our language as well as to have the right hammer to reason about our programs.
From practical side, we need to have the way of taming computational effects
in our programs. That is the point where we would like to introduce to unfamiliar
readers a notion of algebraic effects.

@@

    \noindent
    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
    Universality means that for every algebra $ \beta : F B \longrightarrow B $
    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
    extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram
    below commutes:

@@

    Let $L$, $M$ be models of a theory $T$. $ \mathVar{T-homomorphism}$
    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
    $op_{i}$ in $T$ the following diagram commutes:

@@

    \begin{lemma}{Free models form monads}
    \end{lemma}
    \begin{proof}
        Let $M$ be a free model of an algebraic theory $T$ in category $\mathcal{C}$. We
        have an endofunctor $ F_{T}$, which takes objects into free model and
        maps to unique homomorphisms for which the following diagram commutes:

@@

        \noindent
        which immediately gives us $\eta$ natural transformation for a monad.
        We can now define $\mu{}$ for a monad as a unique morphism for which
        the diagram commutes:

        \begin{center}
        \begin{tikzcd}[row sep=huge, column sep=large]
            F_{T} (X) \arrow[r, "\eta_{F_{T} (X)}"] \arrow[rd, "id"'] & F_{T} (F_{T} (X)) \arrow[d, "\mu{}"] \\
                                                         & F_{T} (Y)
        \end{tikzcd}
        \end{center}

        \noindent
        $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
        From any object we have a unique map sending an object into a model,
        which is $\eta_{F_{T} (X)}$. From free property of our model we get
        a unique map $\mu{}$ such that the diagram commutes, therefore we have
        a monad $(F_{T}, \eta{}, \mu{})$.

@@

    \begin{definition}
        \textit{Initial algebra} is an initial object in \textbf{Alg}, the category of algebras
        and homomorphisms.
    \end{definition}

    \begin{lemma}
        Free algebra over an initial object is initial algebra. Dually,
        cofree coalgebra over a final object is final coalgebra.
    \end{lemma}
    \begin{proof}
        Follows immediately after putting initial object as generator in the
        diagram for free algebra.
    \end{proof}

    Due to close connections presented above of initial algebras and free
    algebras, from which algebraic effects are arising, we could not spare
    a few words a about induction and coinduction, where the latter may give
    us more intuition about coalgebraic effects.

@@

    Here is a table that summarizes difference between these two methods of reasoning:

@@

    From categorical standpoint, coinduction is formed over a final coalgebra,
    where corecursion is the mediator between any coalgebra into a final one,
    and coinductive proof principle corresponds to the uniqueness of the mediator.
    Recall that coalgebraic effects arises from a particular type of a final
    coalgebra, namely, cofree coalgebra.

    \noindent
    TODO Not sure which paper to cite here for a deeper dive into the upper
    topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
    by Adámek, Milius and S. Moss from which I have studied is no longer available
    on the internet, as it was a draft under construction.

@@

    \noindent
    where $\Lambda^{0} \subseteq \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
    meaning, terms without free variables. Let's now define an inductive,
    and coinductive predicate. Relation
    $\Downarrow{} \subseteq \Lambda^{0} \times \Lambda^{0}$ (convergence) for
    call-by-value $\mathVar{\lambda-calculus}$ is defined as follows:

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
    effects are implemented through generators~\cite{one-shot}, using built-in
    feature of sending value when doing \textit{yield} operation. Resumptions in
    handlers that are sending value are one-shot and tail-recursive.

@@

    From theoretical side, research on comodels and its relation to state
    dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
    the runners of comodels to T. Uustalu~\cite{runners-uustalu}.

@@

It can be seen, that issues lie with the fact, that interacting with coalgebras
may lead to change of their internal state system. Examples include change of
state in NFA, taking the next element in an infinite stream or a closing file.
Finalization must occur after initialization, thus we conclude that resumption
should be called at least once. However, modifying state twice the same way can
also lead to unwanted behaviour, as it was seen in
Section~\ref{sec:problem-statement}, therefore it should be invoked exactly
once.

@@

    Where algebraic theory is a name for signature of coeffects that we
    handle, with initial coalgebra carrier holding \verb!<Initial state>! value.
    Handler is syntactically same as in the case of standard algebraic effects,
    however, the parameter passed to cohandler is always a pair of
    current configuration and actual argument passed by user to effect.

@@

    The first example we are going to show is file handling. As per
    Section~\ref{sec:problem-statement}, we do not want to expose file handle
    to the user, thus we hide it behind carrier of FileIO coalgebraic theory,
    for which the signature is composed of Open, Write and Close operations.

@@

    In the above code, we initialize the coalgebra carrier with filename that
    we want to handle, we open a file, write to it once and then close it. Notice
    that file handle is not present in the user code, but only accessible in
    the runner of FileIO as a first element of a pair.

@@

    This example shows how one can implement NFA using coeffects. We abstract
    over internal state of the automata, and based on letters from the alphabet,
    we change automata configuration and return whether word is accepted or not.
    The code below returns whether a given NFA accepts the word "abba".

\begin{verbatim}
    cohandle Automata using 0 at
        let _ <- observe NFA "a" in
        let _ <- observe NFA "b" in
        let _ <- observe NFA "b" in
        observe NFA "a"
    through {
        NFA p r ->
            let state <- return fst p in
            let letter <- return snd p in
            <code transitions here> |
        return x -> return x
    }
\end{verbatim}

@@

    As of this day, two implementations are available, one based on the curried
    translation from Appel~\cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit
    stacks~\cite{handlers-cps-journal}. More details can be found in
    Section~\ref{chapter:implementation}. All commands are available within the
    \verb!src! directory. Please note, however, that Apple-based version is no
    longer supported, and is left as a reference.

@@

    \begin{itemize}
        \item Run tests: \verb!make tests!
        \item Run code linter: \verb!make lint!
        \item Validate Freak code from \verb!programs! dir: \verb!make lint!
        \item Run all above commands: \verb!make check!
    \end{itemize}

@@

    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
    `string' a string value, $V \oplus W$ and $V \approx W$ are respectively
    binary and relational operators, where we support basic arithmetic and
    comparison operations. \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$
    is a standard branching statement. The other constructs are just as in
    Links~\cite{handlers-cps-journal}, with slight syntax modifications.
    Actual programs in Freak can be found in Section~\ref{sec:examples}.

@@

    Semantic of Freak is heavily based on Links language~\cite{handlers-cps-journal},
    for which the source language's dynamics have been described extensively by
    providing small-step operational semantics, continuation passing style
    transformation~\cite{handlers-cps} as well as abstract
    machine~\cite{liberating-effects}, which was proved to coincide with CPS
    translation. That being said, Freak introduces new basic constructs to the
    language, for which we shall define the semantics.

@@

    TODO it's not implemented yet. Be more clear what's happening here

>>>>

Algebraic effects, while not a new concept, have received a lot of
attention in recent years~\cite{effects-bibliography}, both from the theoretical
and practical side.

They give the developer necessary tools for declarative programming with
computational effects on a high level of abstraction, by defining
effects as an API\@ in the code. At the same time, algebraic effects preserve
a lot of flexibility and have a strong theoretical foundations.

@@

In particular, multi-shot resumptions in effect handlers and the possibility to
drop the resumption allow one to express fairly complex logic in a concise way,
however, they also give rise to issues that would not occur in standard
control flow, especially when composing various effects together. Unwanted
interaction with external resources multiple times or not invoking finalisation
code are cases that may occur while using algebraic effects and handlers. Let's
consider the following example:

\vbox{\begin{verbatim}
    handle
        let fh <- do Open "praise.txt" in
        let c <- do Choice () in
        let text <- if c then return "Guy Fieri" else return "is cool" in
        let _ <- do Write (fh, text) in
        do Close fh
    with {
        Choose _ r ->
            let t <- r 1 in
            let f <- r 0 in
            return (t, f) |
        return x -> return x
    }
\end{verbatim}
}

@@

Other branch of languages theory, that we shall investigate in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed, most notably, type systems. Thanks to strong
and static type systems along with their implementations, we have solid tools to
work efficiently on functions that are pure. That being said, we claim that the
core complexity of programs comes from side effects, or more generally, computational
effects, which we cannot avoid in writing anything useful.

@@

% TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource

Not only we would like to bring back effects to our type
system~\cite{type-and-effect}, but also do it in a way that is composable. This
is where algebraic effects come to the rescue. From theoretical point of view,
we need to develop equational theory about effects to assert correctness of
the language, as well as to have the right hammer to reason about our programs.
From practical side, we need to have \textit{the} way of taming computational effects
in our programs. That is the point where we would like to introduce to unfamiliar
readers a notion of algebraic effects.

@@

    \noindent
    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
    Universality means that for every algebra $ \beta : F B \longrightarrow B $
    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
    extending $f$, i.e.\ a unique morphism in $ \mathcal{C}$ for which the diagram
    below commutes:

@@

    Let $L$, $M$ be models of a theory $T$. $ \mathVar{T-homomorphism}$
    $\varphi : L \rightarrow M$ is a map, such that, for every operation symbol
    $op_{i}$ in $T$ the following diagram commutes:

@@

    \begin{lemma}{Free models form monads}
    \end{lemma}
    \begin{proof}
        Let $M$ be a free model of an algebraic theory $T$ in category $\mathcal{C}$. We
        have an endofunctor $ F_{T} : \mathcal{C} \rightarrow \mathcal{C} $, which takes
        objects into free models and maps to unique homomorphisms, for which the following
        diagram commutes:

@@

        \noindent
        which immediately gives us $\eta : 1_{C} \rightarrow F_{T}$ natural
        transformation for a monad.
        We can now define $\mu{} : {F_{T}}^{2} \rightarrow F_{T}$ for a monad
        as a collecton indexed by $X$ of unique morphisms for which
        the diagram commutes:

        \begin{center}
        \begin{tikzcd}[row sep=huge, column sep=large]
            F_{T} (X) \arrow[r, "\eta_{F_{T} (X)}"] \arrow[rd, "id"'] & F_{T} (F_{T} (X)) \arrow[d, "\mu_{F_{T}(X)}"] \\
                                                         & F_{T} (Y)
        \end{tikzcd}
        \end{center}

        \noindent
        $F_{T}$ is an endofunctor, thus $F_{T}(X)$ is an object in $\mathcal{C}$.
        From any object we have a unique map sending an object into a model,
        which is $\eta_{F_{T} (X)}$. From free property of our model we get
        a unique map $\mu{}$ such that the diagram commutes, therefore we have
        a monad $(F_{T}, \eta{}, \mu{})$.

@@

    Due to close connections of initial algebras and free algebras, from which
    algebraic effects are arising, we spare a few words a about
    induction and coinduction, where the latter may give us more intuition
    about coalgebraic effects.

    \begin{definition}
        \textit{Initial algebra} is an initial object in \textbf{Alg}, the category of algebras
        and homomorphisms. Dually, \textit{final coalgebra} is a terminal
        object in \textbf{Alg}
    \end{definition}

    \begin{lemma}
        Free algebra over an initial object is initial algebra.
    \end{lemma}
    \begin{proof}
        Follows immediately after putting initial object as generator in the
        diagram for free algebra. Notice that from duality we have that cofree
        coalgebra over a terminal object is a final coalgebra.
    \end{proof}

@@

    Here is a table that summarizes the differences between these two methods of reasoning:

@@

    From categorical standpoint, coinduction is formed over a final coalgebra,
    where corecursion is the mediator between any coalgebra into a final one,
    and coinductive proof principle corresponds to the uniqueness of the mediator.
    Recall from Chapter~\ref{chapter:background} that coalgebraic effects arises from cofree coalgebra.

    % TODO Not sure which paper to cite here for a deeper dive into the upper
    % topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
    % by Adámek, Milius and S. Moss from which I have studied is no longer available
    % on the internet, as it was a draft under construction.

@@

    \noindent
    We denote $\Lambda^{0} \subseteq \Lambda$ as a set of $\mathVar{closed~\lambda-terms}$,
    meaning, terms without free variables. Let's now define an inductive,
    and coinductive predicate. Relation
    $\Downarrow{} \subseteq \Lambda^{0} \times \Lambda^{0}$ (convergence) for
    call-by-value $\mathVar{\lambda-calculus}$ is defined as follows:

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
    effects are implemented through generators, using built-in feature of
    sending value when doing \textit{yield} operation. Implementations of effects
    based on coroutines~\cite{one-shot}, such as in Python, have one-shot and
    tail-recursive resumptions.

@@

    From theoretical side, research on comodels and its relation to state
    dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
    the runners of comodels from T. Uustalu~\cite{runners-uustalu}.

@@

It can be seen, that issues lie with the fact, that interacting with coalgebras
may lead to change of their internal state system. Examples include change of
state in NFA, taking the next element in an infinite stream or closing a file.
Finalization must occur after initialization, thus we conclude that resumption
should be called at least once. However, modifying state twice the same way can
also lead to unwanted behaviour, as it was seen in
Section~\ref{sec:problem-statement}, therefore it should be invoked exactly
once.

@@

    Where algebraic theory is a name for signature of coeffects that we
    handle, with final coalgebra carrier holding \verb!<Initial state>! value.
    Handler is syntactically same as in the case of standard algebraic effects,
    however, the parameter passed to cohandler is always a pair of
    current configuration and actual argument passed by user to effect.

@@

    The first example we are going to show is file handling. As per
    Section~\ref{sec:problem-statement}, we do not want to expose file handle
    to the user, thus we hide it behind carrier of \textit{FileIO} coalgebraic
    theory, for which the signature is composed of \textit{Open}, \textit{Write}
    and \textit{Close} operations.

@@

    In the above code, we initialize the coalgebra carrier with filename that
    we want to handle, we open a file, write to it once and then close it. Notice
    that file handle is not present in the user code, but only accessible in
    the runner of \textit{FileIO} as a first element of a pair.

@@

    This example shows how one can implement NFA using coeffects. We abstract
    over internal state of the automata, and based on letters from the alphabet,
    we change automata configuration and return whether word is accepted or not.
    The code below returns whether a given NFA accepts the word "Abba".

\begin{verbatim}
    cohandle Automata using 0 at
        let _ <- observe NFA "A" in
        let _ <- observe NFA "b" in
        let _ <- observe NFA "b" in
        observe NFA "a"
    through {
        NFA p r ->
            let state <- return fst p in
            let letter <- return snd p in
            <code transitions here> |
        return x -> return x
    }
\end{verbatim}

@@

    All commands are available within the \verb!src! directory.

@@

    \begin{itemize}
        \item Run tests: \verb!make tests!
        \item Run code linter: \verb!make lint!
        \item Validate Freak code from \verb!programs! dir: \verb!make validate_programs!
        \item Run all above commands: \verb!make check!
    \end{itemize}

@@

    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
    `string' a string value, $V \oplus W$ and $V \approx W$ are respectively
    binary and relational operators, where we support basic arithmetic and
    comparison operations. \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$
    is a standard branching statement. The other constructs are just as in
    Links~\cite{handlers-cps-journal}, with slight syntax modifications.
    Actual programs in Freak can be found in Section~\ref{sec:examples}.
    Derivation of a value for a simple program with function application based
    on CPS calculus can be seen in \verb!derivation! file.

@@

    Semantic of Freak is heavily based on Links language~\cite{handlers-cps-journal},
    for which the source language's dynamics have been described extensively by
    providing small-step operational semantics, continuation passing style
    transformation~\cite{handlers-cps} as well as abstract
    machine~\cite{liberating-effects}, which was proved to coincide with CPS
    translation. That being said, Freak introduces coalgebraic effects and a few
    basic constructs to the language, for which we shall define the semantics.

@@

%     TODO it's not implemented yet. Be more clear what's happening here
